Comparing probabilistic accounts of probability judgments

### 2021-11-05 paper-rmd/paper-papaja.tex
more presubmission updates

\hypertarget{accounting-for-model-complexity}{%
\subsubsection{Accounting for Model Complexity}\label{accounting-for-model-complexity}}

When comparing and selecting models based on their predictive accuracy, it is necessary to correct for the potential for these models to ``overfit'' the data. Typically this correction comes in the form of a ``complexity penalty,'' penalizing models in proportion to their flexibility for accommodating different patterns of data \autocite{gelman.etal2014}. Zhu and colleagues \autocite*{zhu.etal2020} warn that BIC, which penalizes models based solely on the number of parameters in the model, cannot fully account for the differences in the models' complexity.

There are at least three challenges to accounting for model complexity in the comparison of the PT+N and Bayesian Sampler models. First, the models differ not only in the number of parameters but in the domain of those parameters. Zhu and colleagues \autocite*{zhu.etal2020} assume that the Bayesian Sampler model's prior distribution should reflect ignorance or a lack of information. A uniform prior, Beta(1, 1) is the most obvious choice in this case, but theoretical arguments can also be made for Beta(0.5, 0.5) (Jeffrey's prior), Beta(0,0) (Haldene's prior), or perhaps other symmetric beta distributions Beta(\(\beta\),\(\beta\)) with \(\beta \in [0, 1]\) \autocite{jaynes2003}. Via the bridging conditions, they show that assuming \(\beta \in [0, 1]\) restricts the ``noise level'' for the Bayesian Sampler, represented in implied \(d\) under the PT+N model, to {[}0, \(1/3\){]} (approaching \(1/3\) as \(N\) approaches \(\infty\) and \(\beta\) approaches 1), whereas the PT+N model permits noise values between \([0, 1/2]\).

@@

Finally, one potential explanation for the relative weakness of the ``complex'' model variants is Zhu and colleagues' \autocite*{zhu.etal2020} use of ``unpooled'' models, where parameters are estimated independently for each individual participant. In contrast, a comparison of fully pooled variants of the simple and complex models (where a single population parameter is used for all participants) would require adding only one extra parameter to the penalty term. If there is limited heterogeneity across individuals, then adding a parameter for each participant may effectively over-penalize the complex variants relative to the simple variants. Partial pooling is a solution that balances between these extreme approaches, allowing for an accounting of heterogeneity without over-penalizing in cases where heterogeneity is low.

@@

Here, I cast both the Bayesian Sampler and PT+N models into a Bayesian data analysis framework that may permit a more decisive comparison. First, Bayesian data analysis allows issues of model complexity to be addressed through comparisons of model fit based on information criteria, such as the widely-applicable information criterion (\({\text{WAIC}}\)) and Pareto smoothed importance sampling approximate leave-one-out cross validation (\(\text{PSIS-LOO}\)). These approaches measure the ability of a model to capture new unseen data by estimating the expected log predictive density (\(\widehat{\text{elpd}}\)) of the model, or the expected probability of new unseen data. Importantly, these the methods for computing the \(\widehat{\text{elpd}}\) penalize model complexity in a way that directly captures differences in the flexibility of the models, instead of being estimated based solely on the number of distinct parameters \autocite{gelman.etal2014,vehtari.etal2017}. In addition, the Bayesian framework also supports straightforward implementation of hierarchical versions of these models with partial pooling. This allows for information about model parameters to be shared across participants, resulting in a potential reduction in model complexity and a more realistic test of the models.

@@

Both of these models can also be implemented as hierarchical models with partial pooling for the \(d\) and \(d'\) parameters (implicitly, for \(N\) and \(N'\) in the case of the Bayesian Sampler). This should prevent over-penalizing complex versions of the models and potentially improve their overall predictive fit. The hierarchical implementation adds parameters for the population-level \(d\) and \(d'\) as well as a parameter controlling the standard deviation of the distribution for the subject-level effects. For ease of interpretation, the centered parameterization is shown below, although the actual models used a non-centered parameterization to improve sampling efficiency \autocite{papaspiliopoulos.etal2007}. Figure 2 displays the translation of a hierarchical implementation of the Bayesian Sampler model into the Bayesian framework, along with a plate diagram representing the dependencies among parameters.

@@

Finally, I also explored fitting versions of the Bayesian Sampler model that allowed values of \(\beta > 1\). Restricting \(\beta\) to {[}0,1{]} restricts the prior distribution of the Bayesian sampler to the class of ``ignorance priors'' \autocite{zhu.etal2020}. However, it is also possible that people bring informative priors to the probability judgment task. Indeed, Zhu and colleagues (2020) acknowledge there are situations where an informative prior may be warranted \autocite[see e.g.,][]{fennell.baddeley2012}. If \(\beta\) is unrestricted, allowed to fall in the domain \([0, \infty]\) then the Bayesian Sampler model becomes more flexible, allowing for equivalent ``noise'' levels in the same \([0, 1/2]\) range as the PT+N model. That is, through the bridging condition, the implied \(d\) approaches \(1/2\) in the limit as \(N\) and \(\beta\) both approach \(\infty\). Though it would seem a more fundamental change, this same model may also be seen as a version of the PT+N theory that jettisons its constructive account of conditional probability judgment. Thus, fitting this additional unrestricted model allows for a complete comparison of the models along both of their differing dimensions.

@@

Recall, the unrestricted Bayesian Sampler model may also be seen as a version of the PT+N theory that removes the original theory's constructive account of conditional probability judgments. So, comparing the PT+N models to the unrestricted Bayesian Sampler model offers a test of the PT+N's constructive account of conditional probability jugments. Comparing predictions from the unrestricted Bayesian Sampler model and the best-fitting PT+N model, we see that the Bayesian Sampler model better captures these judgments from both experiments (see Table 2). And comparing their prediction specifically for conditional probability judgments, the unrestricted Bayesian Sampler again provides a better fit, especially for the modestly-correlated events of Experiment 2 (Exp 1: response-level \(r\) = .91 vs.~.89; query-level \(r\) = .98 vs.~.97; Exp 2: response-level \(r\) = .72 vs.~.67; query-level \(r\) = .92 vs.~.85) These findings suggest that conditioning is better seen as part of a mental model of the events than as part of the probability judgment process.

Compared to the Bayesian Sampler model, the PT+N model with its constructive account of conditional probability judgments has a similar penalty term estimate when fit to Experiment 1, but has a smaller penalty term when fit to Experiment 2, despite having the same parameterization in terms of \(d\) and \(d'\). Evidently this structural component isn't a complexifying epicycle, instead it is better seen as a prediction that constrains flexibility. However, it is one that ultimately leads to a worse-fitting model.

Perhaps surprisingly, the Bayesian Sampler with unrestricted \(\beta\) actually receives a smaller penalty term than the more ``restricted'' version of the model for the data in Experiment 2. This is at first counterintuitive, but it illustrates how model complexity depends not only on the model and priors, but also the observed data \autocite[see][]{gelman.etal2014}. To illustrate, Gelman and colleagues consider a case where a parameter is constrained to be positive and its value is then estimated from data \autocite*{gelman.etal2014}. If the estimated value is some very large positive number, then the constraint won't have been very informative. But, if the estimated value is very close to zero, then the constraint that the parameter is positive will provide substantial information and the model's penalty term will therefore be smaller. Here, it seems reasonable to conjecture that because the implied \(d\) and \(d'\) estimated for this data under this model are both very near \(1/3\), the restriction results in posterior estimates of the linear parameters that are relatively far from the prior, which can result in a greater penalty.

The dependence of complexity penalties on observed data may strike some as an undesirable feature of model comparison through information criteria. Indeed, it is worth acknowledging that principled model comparison is still an area of active inquiry, with differing perspectives \autocites[e.g.][]{gronau.wagenmakers2019,vehtari.etal2019}.

@@

\caption{\label{tab:table2}Bayesian model comparison results with best scoring model in bold face.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lrrrrrrrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{4}{c}{Experiment 1} & \multicolumn{4}{c}{Experiment 2} \\
\cmidrule(l{3pt}r{3pt}){2-5} \cmidrule(l{3pt}r{3pt}){6-9}
Model & $\widehat{\text{elpd}}_{\text{LOO}}$ & Penalty & $r_{resp}$ & $r_{query}$ & $\widehat{\text{elpd}}_{\text{LOO}}$ & Penalty & $r_{resp}$ & $r_{query}$\\
\midrule
\textbf{Unrestricted BS hier.} & \textbf{1118.7} & \textbf{259.3} & \textbf{0.884} & \textbf{0.964} & \textbf{1978.6} & \textbf{366.4} & \textbf{0.688} & \textbf{0.878}\\
Bayesian Sampler complex hier. & 1088.3 & 269.6 & 0.883 & 0.960 & 1912.3 & 395.0 & 0.675 & 0.852\\
Bayesian Sampler complex & 1087.8 & 264.6 & 0.883 & 0.961 & 1876.9 & 443.4 & 0.679 & 0.848\\
Bayesian Sampler simple & 1045.6 & 253.8 & 0.874 & 0.952 & 1861.2 & 419.1 & 0.667 & 0.831\\
Bayesian Sampler simple hier. & 1039.6 & 259.1 & 0.874 & 0.953 & 1900.7 & 377.8 & 0.667 & 0.839\\
\addlinespace
PT+N complex hier. & 993.0 & 268.9 & 0.867 & 0.946 & 1902.4 & 351.5 & 0.658 & 0.835\\
PT+N complex & 966.1 & 259.8 & 0.864 & 0.941 & 1886.9 & 395.5 & 0.667 & 0.840\\
PT+N simple hier. & 864.0 & 250.0 & 0.839 & 0.919 & 1821.9 & 305.3 & 0.617 & 0.772\\
PT+N simple & 863.5 & 245.5 & 0.838 & 0.918 & 1821.5 & 319.5 & 0.619 & 0.777\\
Relative Freq. & 649.3 & 289.4 & 0.820 & 0.875 & 643.8 & 424.6 & 0.516 & 0.639\\
\bottomrule
\end{tabular}}
\end{table}

@@

By a fair margin, the model best accounting for the experimental data from Zhu and colleagues \autocite*{zhu.etal2020} was a version of the Bayesian Sampler model without restriction on the range of its \(\beta\) parameters. Alternatively, this model can also be seen as a variant of the PT+N model that removes its account of conditional probability judgments. Thus, what these findings indicate most clearly is that the Bayesian Sampler theory provides a superior account of conditional probability judgments in this task. In keeping with the larger theoretical framework of Bayesian cognitive science, the Bayesian Sampler theory assumes that subjective probabilities underlie people's probability judgments, and that conditional probability judgments are produced by Bayesian conditioning occurring in their mental models of the events in question, rather than as arising from the probability judgment process \autocite{chater.etal2020,zhu.etal2020}.

@@

In addition, the implied \(d\) and \(d'\) noise parameters also varied across individuals. Of note, some participant's implied \(d\) and \(d'\) parameters \emph{were} consistent with the class of ignorance priors, even in Experiment 2 (see supplemental materials). Further research should explore this heterogeneity within and across individuals. Unfortunately, pinning down specific components of the Bayesian Sampler model is a challenge, as the \(\beta\) and \(N\) parameters are not uniquely identifiable from judgment data of the sort examined here.

>>>>

\hypertarget{predictive-generalization-and-model-complexity}{%
\subsubsection{Predictive generalization and model complexity}\label{predictive-generalization-and-model-complexity}}

A chief goal of scientific models is to make accurate predictions about future observations. When comparing and selecting models based on their predictive accuracy, it is crucial to consider how the models will perform on new, as-yet-unseen data. The traditional approach to this issue is to compute the fit of the model to the observed data, but to then correct for the potential for these models to ``overfit'' these data. Typically this correction comes in the form of a ``complexity penalty,'' penalizing models in proportion to their flexibility for accommodating different patterns of data \autocite{gelman.etal2014}. Zhu and colleagues \autocite*{zhu.etal2020} warn that BIC, which penalizes models based solely on the number of parameters in the model, cannot fully account for the differences in the competing models' complexity \autocite[also see][]{piantadosi2018}.

There are at least three challenges to accounting for model complexity in the comparison of the PT+N and Bayesian Sampler models. First, the models differ not only in the number of parameters but in the domain of those parameters. Zhu and colleagues \autocite*{zhu.etal2020} assume that the Bayesian Sampler model's prior distribution should reflect ignorance or a lack of information. A uniform prior, Beta(1, 1) is the most obvious choice in this case, but theoretical arguments can also be made for Beta(0.5, 0.5) (Jeffrey's prior), Beta(0,0) (Haldene's prior), or perhaps other symmetric beta distributions Beta(\(\beta\),\(\beta\)) with \(\beta \in [0, 1]\) \autocite{jaynes2003}. Via the bridging conditions, they show that assuming \(\beta \in [0, 1]\) restricts the ``noise level'' for the Bayesian Sampler, represented in implied \(d\) under the PT+N model, to {[}0, \(1/3\){]} (approaching \(1/3\) as \(N\) approaches 1), whereas the PT+N model permits noise values between \([0, 1/2]\).

@@

Third, one potential explanation for the relative weakness of the ``complex'' model variants is Zhu and colleagues' \autocite*{zhu.etal2020} use of ``unpooled'' models, where parameters are estimated independently for each individual participant. In contrast, a comparison of fully pooled variants of the simple and complex models (where a single population parameter is used for all participants) would require adding only one extra parameter to the penalty term. If there is limited heterogeneity across individuals, then adding a parameter for each participant may effectively over-penalize the complex variants relative to the simple variants. Partial pooling is a solution that balances between these extreme approaches, allowing for an accounting of heterogeneity without over-penalizing in cases where heterogeneity is low.

Finally, it is worth recognizing that formal measures of model complexity cannot be expected to perfectly track notions of simplicity or elegance in scientific explanation \autocites[for some related discussions, see][]{kuhn1977,sober2002,piantadosi2018}. For instance, even if the PT+N model's account of conditional probability judgments constrains its flexibility empirically, it seems clear this added component makes it more complex as a putative scientific explanation.

@@

Here, I cast both the Bayesian Sampler and PT+N models into a Bayesian data analysis framework that may permit a more decisive comparison. First, Bayesian data analysis allows issues of model complexity to be addressed through comparisons of model fit based newer information criteria, such as the widely-applicable information criterion (\({\text{WAIC}}\)) and Pareto smoothed importance sampling approximate leave-one-out cross validation (\(\text{PSIS-LOO}\)). In particular, rather than estimating model fit and then penalizing for model complexity, \(\text{PSIS-LOO}\) estimates out-of-sample prediction performance directly by estimating the expected log predictive density (\(\widehat{\text{elpd}}\)) of the model, or the expected probability of new unseen data \autocite{gelman.etal2014,vehtari.etal2017}. From these calculations, an estimate of model complexity (\(\hat{p}_{\text{LOO}}\)) can also be derived.

In addition, the Bayesian framework also supports straightforward implementation of hierarchical versions of these models with partial pooling. This allows for information about model parameters to be shared across participants, resulting in a potential improvements out-of-sample prediction, reductions in model complexity, and a more realistic test of the models.

@@

Both of these models can also be implemented as hierarchical models with partial pooling for the \(d\) and \(d'\) parameters (implicitly, for \(N\) and \(N'\) in the case of the Bayesian Sampler). This partial pooling can help to regularize parameter estimates and improve out-of-sample predictive performance. In addition, partial pooling effectively reduces model complexity, and could support more realistic comparison between the ``simple'' and ``complex'' variants of the models.

The hierarchical implementation adds parameters for the population-level \(d\) and \(d'\) as well as a parameter controlling the standard deviation of the distribution for the subject-level effects. For ease of interpretation, the centered parameterization is shown below, although the actual models used a non-centered parameterization to improve sampling efficiency \autocite{papaspiliopoulos.etal2007}. Figure 2 displays the translation of a hierarchical implementation of the Bayesian Sampler model into the Bayesian framework, along with a plate diagram representing the dependencies among parameters.

@@

Finally, I also explored fitting versions of the Bayesian Sampler model that allowed values of \(\beta > 1\). Restricting \(\beta\) to {[}0,1{]} restricts the prior distribution of the Bayesian sampler to the class of ``ignorance priors'' \autocite{zhu.etal2020}. However, it is also possible that people bring informative priors to the probability judgment task. Indeed, Zhu and colleagues (2020) acknowledge there are situations where an informative prior may be warranted \autocite[see e.g.,][]{fennell.baddeley2012}. If \(\beta\) is unrestricted, allowed to fall in the domain \([0, \infty]\) then the Bayesian Sampler model becomes more flexible, allowing for equivalent ``noise'' levels in the same \([0, 1/2]\) range as the PT+N model. That is, through the bridging condition, the implied \(d\) approaches \(1/2\) in the limit as \(N \to 1\) and \(\beta \to \infty\). Though it would seem a more fundamental change, this same model may also be seen as a version of the PT+N theory that jettisons its constructive account of conditional probability judgment. Thus, fitting this additional unrestricted model allows for a complete comparison of the models along both of their differing dimensions.

@@

Recall, the unrestricted Bayesian Sampler model may also be seen as a version of the PT+N theory excising its constructive account of conditional probability judgments. So, comparing the PT+N models to the unrestricted Bayesian Sampler model offers a test of the PT+N's constructive account of conditional probability jugments. Comparing predictions from the unrestricted Bayesian Sampler model and the best-fitting PT+N model, we see that the Bayesian Sampler model better captures these judgments from both experiments (see Table 2). And comparing their predictions specifically for conditional probability judgments, the unrestricted Bayesian Sampler again provides a better fit, especially for the modestly-correlated events of Experiment 2 (Exp 1: response-level \(r\) = .91 vs.~.89 (\emph{p} \textless{} .05); query-level \(r\) = .98 vs.~.97; Exp 2: response-level \(r\) = .72 vs.~.67 (\emph{p} \textless{} .01); query-level \(r\) = .92 vs.~.85) These findings suggest that conditioning is better seen as part of a mental model of the events than as part of the probability judgment process.

Table 2 also presents estimates of \(\hat{p}_{\text{LOO}}\), the estimate of the effective number of parameters for each model. Compared to the Bayesian Sampler model, the PT+N model with its constructive account of conditional probability judgments has a similar penalty term estimate when fit to Experiment 1, but has a smaller penalty term when fit to Experiment 2, despite having the same parameterization in terms of \(d\) and \(d'\). Although its special treatment of conditional probability judgments makes it more complex as a putative scientific explanation, this structural component appears to actually constrain its predictive flexibility. However, this constraint leads to a worse-fitting model.

Perhaps surprisingly, the Bayesian Sampler with unrestricted \(\beta\) actually receives a smaller penalty term than the more ``restricted'' version of the model for the data in Experiment 2. This is at first counterintuitive. However, model complexity depends not only on the model and priors, but also the observed data \autocite[see][]{gelman.etal2014}. To illustrate, Gelman and colleagues consider a case where a parameter is constrained to be positive and its value is then estimated from data \autocite*{gelman.etal2014}. If the estimated value is some very large positive number, then the constraint won't have been very informative. But, if the estimated value is very close to zero, then the constraint that the parameter is positive will provide substantial information and the model's penalty term will therefore be smaller. Here, it seems reasonable to conjecture that because the implied \(d\) and \(d'\) estimated from Experiment 2's data are both very near \(1/3\) under this model, the restriction results in posterior estimates of the linear parameters that are relatively far from the prior, which can result in a greater penalty.

The dependence of complexity penalties on observed data may strike some as an undesirable feature of model comparison through information criteria like \(\text{PSIS-LOO}\). Indeed, it is worth acknowledging that principled model comparison is still an area of active inquiry, with differing perspectives \autocites[e.g.][]{gronau.wagenmakers2019,vehtari.etal2019}. Fortunately, in this case, comparisons between the models do not rest solely on differences in estimated model complexity.

@@

\caption{\label{tab:table2}Bayesian model comparison results with best scoring model in bold face.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lrrrrrrrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{4}{c}{Experiment 1} & \multicolumn{4}{c}{Experiment 2} \\
\cmidrule(l{3pt}r{3pt}){2-5} \cmidrule(l{3pt}r{3pt}){6-9}
Model & $\widehat{\text{elpd}}_{\text{LOO}}$ & $\hat{p}_{\text{LOO}}$ & $r_{resp}$ & $r_{query}$ & $\widehat{\text{elpd}}_{\text{LOO}}$ & $\hat{p}_{\text{LOO}}$ & $r_{resp}$ & $r_{query}$\\
\midrule
\textbf{Unrestricted BS hier.} & \textbf{1118.7} & \textbf{259.3} & \textbf{0.884} & \textbf{0.964} & \textbf{1978.6} & \textbf{366.4} & \textbf{0.688} & \textbf{0.878}\\
Bayesian Sampler complex hier. & 1088.3 & 269.6 & 0.883 & 0.960 & 1912.3 & 395.0 & 0.675 & 0.852\\
Bayesian Sampler complex & 1087.8 & 264.6 & 0.883 & 0.961 & 1876.9 & 443.4 & 0.679 & 0.848\\
Bayesian Sampler simple & 1045.6 & 253.8 & 0.874 & 0.952 & 1861.2 & 419.1 & 0.667 & 0.831\\
Bayesian Sampler simple hier. & 1039.6 & 259.1 & 0.874 & 0.953 & 1900.7 & 377.8 & 0.667 & 0.839\\
\addlinespace
PT+N complex hier. & 993.0 & 268.9 & 0.867 & 0.946 & 1902.4 & 351.5 & 0.658 & 0.835\\
PT+N complex & 966.1 & 259.8 & 0.864 & 0.941 & 1886.9 & 395.5 & 0.667 & 0.840\\
PT+N simple hier. & 864.0 & 250.0 & 0.839 & 0.919 & 1821.9 & 305.3 & 0.617 & 0.772\\
PT+N simple & 863.5 & 245.5 & 0.838 & 0.918 & 1821.5 & 319.5 & 0.619 & 0.777\\
Relative Freq. & 649.3 & 289.4 & 0.820 & 0.875 & 643.8 & 424.6 & 0.516 & 0.639\\
\bottomrule
\end{tabular}}
\end{table}

@@

By a fair margin, the models best accounting for the experimental data from Zhu and colleagues \autocite*{zhu.etal2020} were implementations of the Bayesian Sampler theory, with the best single model being a version of the Bayesian Sampler model without restriction on the range of its \(\beta\) parameters. Alternatively, this model can also be seen as a variant of the PT+N model that removes its account of conditional probability judgments. Thus, what these findings indicate most clearly is that the Bayesian Sampler theory provides a superior account of conditional probability judgments in this task. In keeping with the larger theoretical framework of Bayesian cognitive science, the Bayesian Sampler theory assumes that subjective probabilities underlie people's probability judgments, and that conditional probability judgments are produced by Bayesian conditioning occurring in their mental models of the events in question, rather than as arising from the probability judgment process \autocite{chater.etal2020,zhu.etal2020}.

@@

In addition, the implied \(d\) and \(d'\) noise parameters also varied across individuals. Of note, some participants' implied \(d\) and \(d'\) parameters \emph{were} consistent with the class of ignorance priors, even in Experiment 2 (see supplemental materials). Further research should explore this heterogeneity within and across individuals. Unfortunately, pinning down specific components of the Bayesian Sampler model is a challenge, as the \(\beta\) and \(N\) parameters are not uniquely identifiable from judgment data of the sort examined here.
